---
title: "rz_snarl_qpcr_comparison"
author: "Cob Staines"
format: html
editor: visual
---

# RZ-SNARL qPCR Comparison

**Objective**: Quantify diferences in Bd qPCR quantities between extraction and qPCR methods between RZ and SNARL labs.


## Setup
Load packages, set up working directory, establish database connection
```{r setup}
#| code-fold: true
#| code-summary: "Show Setup code"
#| results: hide
librarian::shelf(tidyverse, dbplyr, here, janitor, rio, ggplot2, knitr, kableExtra, RPostgres, DBI, RIBBiTR-BII/ribbitrrr)

wddir = here(Sys.getenv("data_dir"), "bd_qpcr_results", "rz_snarl_qpcr_comparison")

# establish database connection
dbcon = hopToDB("railway", hopReg = FALSE)

```

## Load data
- Pull in experiment data from data file: `qpcr_compare_data_260904.xlsx`
- Pull in ITS1 Bd results from both labs from the RIBBiTR database (source of within-lab replicates)
```{r load}
# load data
snarl_raw = import(here(wddir, "qpcr_compare_data_260904.xlsx"),
                   na = c("NA", "NaN"))

# results from database
db_bd = tbl(dbcon, Id("survey_data", "bd_qpcr_results"))
db_sample = tbl(dbcon, Id("survey_data", "sample"))
db_capture = tbl(dbcon, Id("survey_data", "capture"))

# filter before counting replicates, so replicate counts reflect usable results only
bd_results = db_bd %>%
  inner_join(db_sample, by = "sample_id") %>%
  inner_join(db_capture, by = "capture_id") %>%
  filter(ipc_pass | is.na(ipc_pass),
         standard_target_type == "ITS1",
         !is.na(bd_its1_copies_per_swab),
         qpcr_lab %in% c("snarl", "rz_lab")) %>%
  select(result_id,
         sample_id,
         capture_id,
         sample_name_bd,
         bd_cycle_quant,
         bd_target_quant,
         bd_its1_copies_per_swab,
         extraction_plate_name,
         extraction_lab,
         qpcr_plate_name,
         qpcr_lab,
         extraction_kit,
         taxon_capture,
         comments_qpcr) %>%
  collect()

dbDisconnect(dbcon)
```

## Clean & Pivot
Clean column names, then pivot Bd load values wide by Frog ID
```{r clean}
#| code-fold: true
#| code-summary: "Show Clean & Pivot code"
#| 
snarl_clean = snarl_raw %>%
  clean_names() %>%
  rename(extraction_lab = lab_extract,
         extraction_method = extract_method,
         qpcr_lab = lab_qpcr,
         bd_quant_cycle = quant_cycle,
         bd_start_quant = start_quant,
         frog_id = frog_num) %>%
  mutate(extraction_lab = recode_values(extraction_lab,
                                        "pitt" ~ "rz",
                                        "snarl" ~ "snarl",
                                        default = NA),
         collect_lab = recode_values(collect_lab,
                                     "pitt" ~ "rz",
                                     "snarl" ~ "snarl",
                                     default = NA),
         qpcr_lab = recode_values(qpcr_lab,
                                  "pitt" ~ "rz",
                                  "snarl" ~ "snarl",
                                  default = NA),
         frog_id = as.character(frog_id),
         src = "snarl")

# catch unexpected lab values silently recoded to NA
stopifnot(!anyNA(snarl_clean$extraction_lab),
          !anyNA(snarl_clean$collect_lab),
          !anyNA(snarl_clean$qpcr_lab))

snarl_wide = snarl_clean %>%
  pivot_wider(id_cols = c("frog_id", "collect_lab"),
              names_from = c("extraction_method", "qpcr_lab"),
              values_from = "bd_load")

# database results, excluding experiment swabs so they are not double-counted as replicates
db_clean = bd_results %>%
  filter(!sample_name_bd %in% snarl_clean$bd_swab_id) %>%
  rename(bd_swab_id = sample_name_bd,
         bd_quant_cycle = bd_cycle_quant,
         bd_start_quant = bd_target_quant,
         bd_load = bd_its1_copies_per_swab,
         notes = comments_qpcr) %>%
  mutate(extraction_lab = recode_values(extraction_lab,
                                        "rz_lab" ~ "rz",
                                        default = extraction_lab),
         qpcr_lab = recode_values(qpcr_lab,
                                  "rz_lab" ~ "rz",
                                  default = qpcr_lab),
         extraction_method = recode_values(extraction_kit,
                                           "qiagen_dneasy" ~ "qiagen",
                                           default = extraction_kit),
         collect_lab = NA_character_,  # not recorded in db; not used for replicates
         frog_id = capture_id,
         src = "db",
         species_capture = str_to_sentence(gsub("_", " ", taxon_capture)))

# qPCR replicates: same extract (swab + extraction plate) run more than once within a qPCR lab
qpcr_replicates_clean = db_clean %>%
  group_by(bd_swab_id, extraction_plate_name, qpcr_lab) %>%
  filter(n() > 1) %>%
  mutate(replicate_type = "qpcr",
         replicate_group = paste(bd_swab_id, extraction_plate_name, qpcr_lab, sep = "_"),
         replicate_id = row_number()) %>%
  ungroup()

# swab replicates: multiple swabs from the same capture, through the same extraction & qPCR lab
swab_replicates_clean = db_clean %>%
  # keep one result per swab (first qPCR plate), so qPCR re-runs are not counted as swab replicates
  arrange(qpcr_plate_name) %>%
  group_by(bd_swab_id, extraction_lab, extraction_method, qpcr_lab) %>%
  slice_head(n = 1) %>%
  group_by(capture_id, extraction_lab, extraction_method, qpcr_lab) %>%
  filter(n() > 1) %>%
  mutate(replicate_type = "swab",
         replicate_group = paste(capture_id, extraction_lab, qpcr_lab, sep = "_"),
         replicate_id = row_number()) %>%
  ungroup()

replicates_clean = bind_rows(qpcr_replicates_clean,
                             swab_replicates_clean) %>%
  select(all_of(colnames(snarl_clean)),
         replicate_type,
         replicate_group,
         replicate_id)

replicates_wide = replicates_clean %>%
  pivot_wider(id_cols = c("replicate_type", "replicate_group", "frog_id",
                          "extraction_method", "qpcr_lab"),
              names_from = "replicate_id",
              values_from = "bd_load")

kable(snarl_wide, digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = TRUE,
                font_size = 12,
                fixed_thead = TRUE) %>%
  scroll_box(height = "400px")

```

## Data Summary
Pairwise comparisons of Bd load across extraction method (Qiagen = RZ, PrepMan = SNARL) and qPCR lab. Panels A & B share an extract (qPCR effect only); panels C–E compare different swabs, so they also include swab-to-swab variability.
```{r summary}
#| code-fold: true
#| code-summary: "Show Data Summary code"
#| fig-width: 5
#| fig-height: 5
#| 
nice <- c(qiagen_rz = "Qiagen-RZ", qiagen_snarl = "Qiagen-SNARL",
          prepman_rz = "PrepMan-RZ", prepman_snarl = "PrepMan-SNARL")

# One row per panel; x/y orientation is set here
pairs <- tribble(
  ~letter, ~xvar,      ~yvar,
  "A",     "qiagen_rz",    "qiagen_snarl",
  "B",     "prepman_rz", "prepman_snarl",
  "C",     "qiagen_rz",    "prepman_rz",
  "D",     "qiagen_snarl", "prepman_snarl",
  "E",     "qiagen_rz",    "prepman_snarl"   # different swabs: includes swab-to-swab variability
)

# Stack all pairs into one long data frame
plot_df <- pmap_dfr(pairs, function(letter, xvar, yvar) {
  xv <- snarl_wide[[xvar]]
  yv <- snarl_wide[[yvar]]
  tibble(letter = letter, xvar = xvar, yvar = yvar, x = xv, y = yv)
}) %>%
  filter(!is.na(x), !is.na(y)) %>%
  mutate(xvar = factor(nice[xvar], levels = nice[1:3]),
         yvar = factor(nice[yvar], levels = nice[2:4]))

# Per-panel stats (on raw loads)
stats_df <- plot_df %>%
  group_by(letter, xvar, yvar) %>%
  summarise(n = n(), n_pos = sum(x > 0 & y > 0), .groups = "drop") %>%
  mutate(label = paste0("n = ", n, "\nboth > 0: ", n_pos))

ggplot(plot_df, aes(log10(x + 1), log10(y + 1))) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.6) +
  # geom_smooth(method = "lm") +
  geom_text(data = stats_df, aes(x = -Inf, y = Inf, label = letter),
            hjust = -0.6, vjust = 1.4, fontface = "bold", size = 5,
            inherit.aes = FALSE) +
  geom_text(data = stats_df, aes(x = Inf, y = -Inf, label = label),
            hjust = 1.05, vjust = -0.3, size = 2.75, lineheight = 0.9,
            inherit.aes = FALSE) +
  facet_grid(xvar ~ yvar, switch = "both") +
  coord_fixed() +
  labs(x = "log10(Bd load + 1)  [extraction method – qPCR lab]",
       y = "log10(Bd load + 1)  [extraction method – qPCR lab]") +
  theme_bw() +
  theme(strip.placement = "outside",
        strip.background = element_blank(),
        strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank())

```

## Replicate Variability
Within-lab replicate pairs by qPCR lab and replicate type, to assess whether replicate variability is comparable between labs (i.e. whether replicates can be pooled across labs).
```{r replicates}
#| code-fold: true
#| code-summary: "Show Replicate Variability code"
#| fig-width: 6
#| fig-height: 6

# All within-group replicate pairs (i < j)
replicate_long = replicates_clean %>%
  mutate(pipeline = paste(extraction_method, qpcr_lab, sep = "_")) %>%
  select(replicate_type, replicate_group, pipeline, qpcr_lab, replicate_id, bd_load)

replicate_pairs = replicate_long %>%
  inner_join(replicate_long,
             by = c("replicate_type", "replicate_group", "pipeline", "qpcr_lab"),
             suffix = c("_x", "_y"),
             relationship = "many-to-many") %>%
  filter(replicate_id_x < replicate_id_y) %>%
  transmute(replicate_type, qpcr_lab, pipeline,
            x = bd_load_x,
            y = bd_load_y)

rep_plot_df = replicate_pairs %>%
  mutate(qpcr_lab = recode_values(qpcr_lab,
                                  "rz" ~ "RZ qPCR",
                                  "snarl" ~ "SNARL qPCR",
                                  default = qpcr_lab),
         replicate_type = recode_values(replicate_type,
                                        "qpcr" ~ "qPCR replicates",
                                        "swab" ~ "Swab replicates",
                                        default = replicate_type))

rep_stats_df = rep_plot_df %>%
  group_by(qpcr_lab, replicate_type) %>%
  summarise(n = n(), n_pos = sum(x > 0 & y > 0), .groups = "drop") %>%
  mutate(label = paste0("n = ", n, "\nboth > 0: ", n_pos))

ggplot(rep_plot_df, aes(log10(x + 1), log10(y + 1))) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.6) +
  geom_text(data = rep_stats_df, aes(x = Inf, y = -Inf, label = label),
            hjust = 1.05, vjust = -0.3, size = 2.75, lineheight = 0.9,
            inherit.aes = FALSE) +
  facet_grid(qpcr_lab ~ replicate_type) +
  coord_fixed() +
  labs(x = "log10(Bd load + 1)  [replicate i]",
       y = "log10(Bd load + 1)  [replicate j]") +
  theme_bw() +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank())

```
