---
title: "rz_snarl_qpcr_comparison"
author: "Cob Staines"
format: html
editor: visual
---

# RZ-SNARL qPCR Comparison

**Objective**: Quantify diferences in Bd qPCR quantities between extraction and qPCR methods between RZ and SNARL labs.


## Setup
Load packages, set up working directory, establish database connection
```{r setup}
#| code-fold: true
#| code-summary: "Show Setup code"
#| results: hide
librarian::shelf(tidyverse, dbplyr, here, janitor, rio, ggplot2, knitr, kableExtra, RPostgres, DBI, RIBBiTR-BII/ribbitrrr)

wddir = here(Sys.getenv("data_dir"), "bd_qpcr_results", "rz_snarl_qpcr_comparison")

# establish database connection
dbcon = hopToDB("railway", hopReg = FALSE)

```

## Load data
- Pull in experiment data from data file: `qpcr_compare_data_260904.xlsx`
- Pull in ITS1 Bd results from both labs from the RIBBiTR database (source of within-lab replicates)
```{r load}
# load data
snarl_raw = import(here(wddir, "qpcr_compare_data_260904.xlsx"),
                   na = c("NA", "NaN"))

# results from database
db_bd = tbl(dbcon, Id("survey_data", "bd_qpcr_results"))
db_sample = tbl(dbcon, Id("survey_data", "sample"))
db_capture = tbl(dbcon, Id("survey_data", "capture"))
db_survey = tbl(dbcon, Id("survey_data", "survey"))
db_visit = tbl(dbcon, Id("survey_data", "visit"))
db_site = tbl(dbcon, Id("survey_data", "site"))

# filter before counting replicates, so replicate counts reflect usable results only
bd_results = db_bd %>%
  inner_join(db_sample, by = "sample_id") %>%
  inner_join(db_capture, by = "capture_id") %>%
  left_join(db_survey %>% select(survey_id, visit_id), by = "survey_id") %>%
  left_join(db_visit %>% select(visit_id, date, site_id, project_id), by = "visit_id") %>%
  left_join(db_site %>% select(site_id, site), by = "site_id") %>%
  filter(ipc_pass | is.na(ipc_pass),
         standard_target_type == "ITS1",
         !is.na(bd_its1_copies_per_swab),
         qpcr_lab %in% c("snarl", "rz_lab")) %>%
  select(result_id,
         sample_id,
         capture_id,
         sample_name_bd,
         bd_cycle_quant,
         bd_target_quant,
         bd_its1_copies_per_swab,
         extraction_plate_name,
         extraction_lab,
         qpcr_plate_name,
         qpcr_lab,
         extraction_kit,
         taxon_capture,
         comments_qpcr,
         # metadata for replicate diagnostics
         sample_type,
         sample_name_conflict,
         swab_type,
         qpcr_well_replicates = replicates,
         extraction_date,
         qpcr_date,
         visit_date = date,
         site,
         project_id) %>%
  collect()

dbDisconnect(dbcon)
```

## Clean & Pivot
Clean column names, then pivot Bd load values wide by Frog ID
```{r clean}
#| code-fold: true
#| code-summary: "Show Clean & Pivot code"
#| 
snarl_clean = snarl_raw %>%
  clean_names() %>%
  rename(extraction_lab = lab_extract,
         extraction_method = extract_method,
         qpcr_lab = lab_qpcr,
         bd_quant_cycle = quant_cycle,
         bd_start_quant = start_quant,
         frog_id = frog_num) %>%
  mutate(extraction_lab = recode_values(extraction_lab,
                                        "pitt" ~ "rz",
                                        "snarl" ~ "snarl",
                                        default = NA),
         collect_lab = recode_values(collect_lab,
                                     "pitt" ~ "rz",
                                     "snarl" ~ "snarl",
                                     default = NA),
         qpcr_lab = recode_values(qpcr_lab,
                                  "pitt" ~ "rz",
                                  "snarl" ~ "snarl",
                                  default = NA),
         frog_id = as.character(frog_id),
         src = "snarl")

# catch unexpected lab values silently recoded to NA
stopifnot(!anyNA(snarl_clean$extraction_lab),
          !anyNA(snarl_clean$collect_lab),
          !anyNA(snarl_clean$qpcr_lab))

snarl_wide = snarl_clean %>%
  pivot_wider(id_cols = c("frog_id", "collect_lab"),
              names_from = c("extraction_method", "qpcr_lab"),
              values_from = "bd_load")

# database results, excluding experiment swabs so they are not double-counted as replicates
db_clean = bd_results %>%
  filter(!sample_name_bd %in% snarl_clean$bd_swab_id) %>%
  rename(bd_swab_id = sample_name_bd,
         bd_quant_cycle = bd_cycle_quant,
         bd_start_quant = bd_target_quant,
         bd_load = bd_its1_copies_per_swab,
         notes = comments_qpcr) %>%
  mutate(extraction_lab = recode_values(extraction_lab,
                                        "rz_lab" ~ "rz",
                                        default = extraction_lab),
         qpcr_lab = recode_values(qpcr_lab,
                                  "rz_lab" ~ "rz",
                                  default = qpcr_lab),
         extraction_method = recode_values(extraction_kit,
                                           "qiagen_dneasy" ~ "qiagen",
                                           default = extraction_kit),
         collect_lab = NA_character_,  # not recorded in db; not used for replicates
         frog_id = capture_id,
         src = "db",
         species_capture = str_to_sentence(gsub("_", " ", taxon_capture)))

# qPCR replicates: same extract (swab + extraction plate) run more than once within a qPCR lab
qpcr_replicates_clean = db_clean %>%
  group_by(bd_swab_id, extraction_plate_name, qpcr_lab) %>%
  filter(n() > 1) %>%
  mutate(replicate_type = "qpcr",
         replicate_group = paste(bd_swab_id, extraction_plate_name, qpcr_lab, sep = "_"),
         replicate_id = row_number()) %>%
  ungroup()

# swab replicates: multiple swabs from the same capture, through the same extraction & qPCR lab
swab_replicates_clean = db_clean %>%
  # keep one result per swab (first qPCR plate), so qPCR re-runs are not counted as swab replicates
  arrange(qpcr_plate_name) %>%
  group_by(bd_swab_id, extraction_lab, extraction_method, qpcr_lab) %>%
  slice_head(n = 1) %>%
  group_by(capture_id, extraction_lab, extraction_method, qpcr_lab) %>%
  filter(n() > 1) %>%
  mutate(replicate_type = "swab",
         replicate_group = paste(capture_id, extraction_lab, qpcr_lab, sep = "_"),
         replicate_id = row_number()) %>%
  ungroup()

replicates_clean = bind_rows(qpcr_replicates_clean,
                             swab_replicates_clean) %>%
  select(all_of(colnames(snarl_clean)),
         qpcr_plate_name,
         project_id,
         replicate_type,
         replicate_group,
         replicate_id)

replicates_wide = replicates_clean %>%
  pivot_wider(id_cols = c("replicate_type", "replicate_group", "frog_id",
                          "extraction_method", "qpcr_lab"),
              names_from = "replicate_id",
              values_from = "bd_load")

kable(snarl_wide, digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = TRUE,
                font_size = 12,
                fixed_thead = TRUE) %>%
  scroll_box(height = "400px")

```

## Data Summary
Pairwise comparisons of Bd load across extraction method (Qiagen = RZ, PrepMan = SNARL) and qPCR lab. Panels A & B share an extract (qPCR effect only); panels C–E compare different swabs, so they also include swab-to-swab variability.
```{r summary}
#| code-fold: true
#| code-summary: "Show Data Summary code"
#| fig-width: 5
#| fig-height: 5
#| 
nice <- c(qiagen_rz = "Qiagen-RZ", qiagen_snarl = "Qiagen-SNARL",
          prepman_rz = "PrepMan-RZ", prepman_snarl = "PrepMan-SNARL")

# One row per panel; x/y orientation is set here
pairs <- tribble(
  ~letter, ~xvar,      ~yvar,
  "A",     "qiagen_rz",    "qiagen_snarl",
  "B",     "prepman_rz", "prepman_snarl",
  "C",     "qiagen_rz",    "prepman_rz",
  "D",     "qiagen_snarl", "prepman_snarl",
  "E",     "qiagen_rz",    "prepman_snarl"   # different swabs: includes swab-to-swab variability
)

# Stack all pairs into one long data frame
plot_df <- pmap_dfr(pairs, function(letter, xvar, yvar) {
  xv <- snarl_wide[[xvar]]
  yv <- snarl_wide[[yvar]]
  tibble(letter = letter, xvar = xvar, yvar = yvar, x = xv, y = yv)
}) %>%
  filter(!is.na(x), !is.na(y)) %>%
  mutate(xvar = factor(nice[xvar], levels = nice[1:3]),
         yvar = factor(nice[yvar], levels = nice[2:4]))

# Per-panel stats (on raw loads)
stats_df <- plot_df %>%
  group_by(letter, xvar, yvar) %>%
  summarise(n = n(), n_pos = sum(x > 0 & y > 0), .groups = "drop") %>%
  mutate(label = paste0("n = ", n, "\nboth > 0: ", n_pos))

ggplot(plot_df, aes(log10(x + 1), log10(y + 1))) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.6) +
  # geom_smooth(method = "lm") +
  geom_text(data = stats_df, aes(x = -Inf, y = Inf, label = letter),
            hjust = -0.6, vjust = 1.4, fontface = "bold", size = 5,
            inherit.aes = FALSE) +
  geom_text(data = stats_df, aes(x = Inf, y = -Inf, label = label),
            hjust = 1.05, vjust = -0.3, size = 2.75, lineheight = 0.9,
            inherit.aes = FALSE) +
  facet_grid(xvar ~ yvar, switch = "both") +
  coord_fixed() +
  labs(x = "log10(Bd load + 1)  [extraction method – qPCR lab]",
       y = "log10(Bd load + 1)  [extraction method – qPCR lab]") +
  theme_bw() +
  theme(strip.placement = "outside",
        strip.background = element_blank(),
        strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank())

```

## Replicate Variability
Within-lab replicate pairs by qPCR lab and replicate type, to assess whether replicate variability is comparable between labs (i.e. whether replicates can be pooled across labs).
```{r replicates}
#| code-fold: true
#| code-summary: "Show Replicate Variability code"
#| fig-width: 6
#| fig-height: 6

# All within-group replicate pairs (i < j)
replicate_long = replicates_clean %>%
  mutate(pipeline = paste(extraction_method, qpcr_lab, sep = "_")) %>%
  select(replicate_type, replicate_group, pipeline, qpcr_lab, project_id, replicate_id, qpcr_plate_name, bd_load)

replicate_pairs = replicate_long %>%
  inner_join(replicate_long,
             by = c("replicate_type", "replicate_group", "pipeline", "qpcr_lab", "project_id"),
             suffix = c("_x", "_y"),
             relationship = "many-to-many") %>%
  filter(replicate_id_x < replicate_id_y) %>%
  transmute(replicate_type, qpcr_lab, pipeline, project_id,
            same_qpcr_plate = qpcr_plate_name_x == qpcr_plate_name_y,
            x = bd_load_x,
            y = bd_load_y)

rep_plot_df = replicate_pairs %>%
  mutate(qpcr_lab = recode_values(qpcr_lab,
                                  "rz" ~ "RZ qPCR",
                                  "snarl" ~ "SNARL qPCR",
                                  default = qpcr_lab),
         replicate_type = recode_values(replicate_type,
                                        "qpcr" ~ "qPCR replicates",
                                        "swab" ~ "Swab replicates",
                                        default = replicate_type))

rep_stats_df = rep_plot_df %>%
  group_by(qpcr_lab, replicate_type) %>%
  summarise(n = n(), n_pos = sum(x > 0 & y > 0), .groups = "drop") %>%
  mutate(label = paste0("n = ", n, "\nboth > 0: ", n_pos))

ggplot(rep_plot_df, aes(log10(x + 1), log10(y + 1))) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.6) +
  geom_text(data = rep_stats_df, aes(x = Inf, y = -Inf, label = label),
            hjust = 1.05, vjust = -0.3, size = 2.75, lineheight = 0.9,
            inherit.aes = FALSE) +
  facet_grid(qpcr_lab ~ replicate_type) +
  coord_fixed() +
  labs(x = "log10(Bd load + 1)  [replicate i]",
       y = "log10(Bd load + 1)  [replicate j]") +
  theme_bw() +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank())

```

## Replicate Diagnostics
Swab replicates appear more variable than the different-swab experiment comparisons (panels C–E). These diagnostics check whether that is real variability, or an artifact of how replicate groups are defined (non-duplicate swabs, name conflicts, batch/plate effects, specific projects or sites).

### Pair agreement
Share of pairs by detection outcome, and SD of the log10 difference among pairs where both are positive. Replicates are split by whether the pair was run on the same qPCR plate (for qPCR replicates, same plate = technical well replicates).
```{r diag-pairs}
#| code-fold: true
#| code-summary: "Show Pair agreement code"

summarise_pairs = function(df) {
  df %>%
    summarise(n_pairs = n(),
              both_zero = mean(x == 0 & y == 0),
              one_zero = mean(xor(x > 0, y > 0)),
              both_pos = mean(x > 0 & y > 0),
              sd_log10_diff_pos = sd((log10(y) - log10(x))[x > 0 & y > 0]),
              .groups = "drop")
}

experiment_agreement = plot_df %>%
  mutate(dataset = paste0("Experiment ", letter, ": ", xvar, " vs ", yvar),
         shared_extract = letter %in% c("A", "B")) %>%
  group_by(dataset, shared_extract) %>%
  summarise_pairs()

replicate_agreement = replicate_pairs %>%
  mutate(dataset = paste0(replicate_type, " replicates: ", pipeline, ", ",
                          if_else(same_qpcr_plate, "same qPCR plate", "different qPCR plates")),
         shared_extract = replicate_type == "qpcr") %>%
  group_by(dataset, shared_extract) %>%
  summarise_pairs()

bind_rows(experiment_agreement, replicate_agreement) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = TRUE,
                font_size = 12)
```

### Swab replicate groups
Per capture × pipeline group: number of swabs, sample types, name conflicts, number of extraction & qPCR plates, and spread of loads.
```{r diag-swab-groups}
#| code-fold: true
#| code-summary: "Show Swab replicate groups code"

swab_groups = swab_replicates_clean %>%
  group_by(qpcr_lab, replicate_group) %>%
  summarise(n_swabs = n(),
            n_sample_types = n_distinct(sample_type),
            name_conflict = any(!is.na(sample_name_conflict)),
            n_extraction_plates = n_distinct(extraction_plate_name),
            n_qpcr_plates = n_distinct(qpcr_plate_name),
            qpcr_days_apart = as.numeric(max(as.Date(qpcr_date)) - min(as.Date(qpcr_date))),
            min_load = min(bd_load),
            max_load = max(bd_load),
            log10_range = log10(max_load + 1) - log10(min_load + 1),
            discordant = min_load == 0 & max_load > 0,
            project_id = first(project_id),
            site = first(site),
            visit_date = first(visit_date),
            .groups = "drop")

# group-level flags by lab
swab_groups %>%
  group_by(qpcr_lab) %>%
  summarise(n_groups = n(),
            more_than_2_swabs = mean(n_swabs > 2),
            multiple_sample_types = mean(n_sample_types > 1),
            name_conflict = mean(name_conflict),
            multiple_extraction_plates = mean(n_extraction_plates > 1),
            multiple_qpcr_plates = mean(n_qpcr_plates > 1),
            median_qpcr_days_apart = median(qpcr_days_apart, na.rm = TRUE),
            discordant = mean(discordant),
            median_log10_range = median(log10_range),
            .groups = "drop") %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = TRUE,
                font_size = 12)

# is disagreement concentrated in specific projects?
swab_groups %>%
  group_by(qpcr_lab, project_id) %>%
  summarise(n_groups = n(),
            discordant = mean(discordant),
            median_log10_range = median(log10_range),
            .groups = "drop") %>%
  arrange(qpcr_lab, desc(n_groups)) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = TRUE,
                font_size = 12)
```

### Largest swab replicate disagreements
Individual swabs from the 15 groups with the largest load range.
```{r diag-swab-worst}
#| code-fold: true
#| code-summary: "Show Largest disagreements code"

worst_groups = swab_groups %>%
  slice_max(log10_range, n = 15, with_ties = FALSE) %>%
  pull(replicate_group)

swab_replicates_clean %>%
  filter(replicate_group %in% worst_groups) %>%
  select(replicate_group, bd_swab_id, sample_type, sample_name_conflict,
         site, visit_date, extraction_plate_name, extraction_date,
         qpcr_plate_name, qpcr_date, qpcr_well_replicates, bd_load, notes) %>%
  arrange(replicate_group, bd_swab_id) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = TRUE,
                font_size = 11) %>%
  scroll_box(height = "400px")
```

### Difference vs. mean
Bland–Altman style: difference in log10(Bd load + 1) between the two members of each pair, against their mean. Lines are smoothed mean absolute differences. If spread shrinks with load and populations/projects follow the same curve, the differences between panels are explained by load rather than population or lab.
```{r diag-bland-altman}
#| code-fold: true
#| code-summary: "Show Difference vs. mean code"
#| fig-width: 9
#| fig-height: 8

# experiment pairs, keeping collecting lab (population) per frog
experiment_pairs = pmap_dfr(pairs, function(letter, xvar, yvar) {
  tibble(letter = letter,
         dataset = paste0("Experiment ", letter, ": ", nice[xvar], " vs ", nice[yvar]),
         collect_lab = snarl_wide$collect_lab,
         x = snarl_wide[[xvar]],
         y = snarl_wide[[yvar]])
}) %>%
  filter(!is.na(x), !is.na(y))

ba_df = bind_rows(
  experiment_pairs %>%
    transmute(dataset,
              different_swabs = !letter %in% c("A", "B"),
              group = recode_values(collect_lab,
                                    "rz" ~ "Collected: RZ (Pittsburgh)",
                                    "snarl" ~ "Collected: SNARL (Sierra Nevada)",
                                    default = collect_lab),
              x, y),
  replicate_pairs %>%
    transmute(dataset = paste0(replicate_type, " replicates: ", nice[pipeline]),
              different_swabs = replicate_type == "swab",
              group = paste("DB project:", project_id),
              x, y)) %>%
  mutate(mean_log = (log10(x + 1) + log10(y + 1)) / 2,
         diff_log = log10(y + 1) - log10(x + 1),
         outcome = case_when(x == 0 & y == 0 ~ "both zero",
                             x == 0 | y == 0 ~ "one zero",
                             TRUE ~ "both positive"))

# all datasets
ggplot(ba_df, aes(mean_log, diff_log, colour = group)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(aes(shape = outcome), alpha = 0.6) +
  geom_smooth(aes(y = abs(diff_log)), method = "loess", se = FALSE, linewidth = 0.7) +
  facet_wrap(~ dataset) +
  scale_shape_manual(values = c("both zero" = 4, "one zero" = 1, "both positive" = 16)) +
  labs(x = "mean log10(Bd load + 1)",
       y = "difference in log10(Bd load + 1)  (line: mean |difference|)",
       colour = NULL, shape = NULL) +
  theme_bw() +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.direction = "vertical")

```

Overlay of the different-swab comparisons only (experiment C–E and swab replicates), to compare noise-vs-load curves directly.
```{r diag-bland-altman-overlay}
#| code-fold: true
#| code-summary: "Show overlay code"
#| fig-width: 7
#| fig-height: 5

ba_df %>%
  filter(different_swabs, outcome != "both zero") %>%
  ggplot(aes(mean_log, abs(diff_log), colour = dataset)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.9) +
  labs(x = "mean log10(Bd load + 1)",
       y = "|difference in log10(Bd load + 1)|",
       colour = NULL) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.direction = "vertical")
```
