---
title: "rz_snarl_qpcr_model1_replicates"
author: "Cob Staines"
format: html
editor: visual
---

# RZ-SNARL qPCR Comparison: Model 1 (Replicates)

**Objective**: Model within-lab noise and detection from replicates. Posterior draws and summaries are exported for use as priors in Model 2 (`rz_snarl_qpcr_model2_experiment.qmd`).

## Setup

Load packages; load and clean experiment and replicate data (`R/data_prep.R`, from data saved by `R/data_import.R`)

```{r setup}
#| code-fold: true
#| code-summary: "Show Setup code"
#| results: hide
librarian::shelf(tidyverse, here, knitr, kableExtra, ggplot2, patchwork, brms, cmdstanr, posterior, bayesplot)

cmdstanr::check_cmdstan_toolchain()   # confirms Stan is found
set_cmdstan_path(Sys.getenv("cmd_stan_path"))

source(here("R", "data_prep.R"))

# fitted models are saved outside the repo
fit_dir = here(wddir, "fits")
dir.create(fit_dir, showWarnings = FALSE)

# set TRUE to refit models instead of loading saved fits
refit = FALSE
```

## Model 1: Replicates
Within-lab noise and detection as a function of load, as a brms hurdle-lognormal model (cmdstanr backend).

- **Latent load**: each replicate group (a capture for swab replicates, an extract for qPCR replicates) has a group-level intercept in `mu`, i.e. its true (log) load.
- **Detection**: `hu` (probability of a zero) has a qPCR lab effect and a group-level intercept. The group-level intercepts of `mu` and `hu` are correlated (shared ID `g`), so the `mu`–`hu` correlation captures load-dependent detection (expected negative: higher load, fewer zeros).
- **Noise**: `sigma` differs by noise type: swab replicates (total noise: swab + extraction + qPCR) and qPCR replicates (total noise between qPCR re-runs of the same extract).

Kept deliberately simple given limited replicates. An earlier version also had a qPCR run (extract × plate) effect and a `sigma` group-level intercept (load-dependent noise); neither was identified (nearly every qPCR run was a single well; 2–3 positives per group), and they caused divergent transitions.

`mu` and `sigma` are on the natural-log scale of Bd load (ITS1 copies per swab); results are converted to log10 for plots.

### Model data
```{r m1-data}
#| code-fold: true
#| code-summary: "Show Model data code"

m1_df = replicates_clean %>%
  mutate(qpcr_lab = factor(qpcr_lab, levels = c("rz", "snarl")),
         noise_type = factor(replicate_type, levels = c("swab", "qpcr")),
         ppc_group = paste(replicate_type, qpcr_lab))

m1_df %>%
  group_by(replicate_type, qpcr_lab) %>%
  summarise(groups = n_distinct(replicate_group),
            results = n(),
            qpcr_plates = n_distinct(qpcr_plate_name),
            detected = mean(bd_load > 0),
            .groups = "drop") %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### Model structure
```{r m1-structure}
#| code-fold: true
#| code-summary: "Show Model structure code"

m1_family = hurdle_lognormal()

m1_formula = bf(
  bd_load ~ 1 +
    (1 | g | replicate_group),

  hu ~ 1 + qpcr_lab +
    (1 | g | replicate_group),

  sigma ~ 0 + noise_type
)

m1_priors = c(
  # mu (ln copies per swab)
  prior(normal(7, 4), class = "Intercept"),

  # hu
  prior(normal(0, 2), class = "Intercept", dpar = "hu"),
  prior(normal(0, 2), class = "b", dpar = "hu"),

  # sigma (log of ln-scale SD)
  prior(normal(0, 1), class = "b", dpar = "sigma"),

  # group-level effects
  prior(exponential(1), class = "sd"),
  prior(lkj(2), class = "cor")
)
```

### Fit
Fit is saved to `fits/` in the data directory and reloaded unless `refit = TRUE`.
```{r m1-fit}
#| code-fold: true
#| code-summary: "Show Fit code"
#| results: hide

m1 = brm(
  formula = m1_formula,
  data = m1_df,
  family = m1_family,
  prior = m1_priors,
  control = list(adapt_delta = 0.98),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 1234,
  cores = 4,
  backend = "cmdstanr",
  file = here(fit_dir, "m1_replicates"),
  file_refit = if (refit) "always" else "on_change"
)
```

### Convergence & summary
```{r m1-summary}
#| code-fold: true
#| code-summary: "Show Summary code"

summary(m1)

# group-level correlation: mu-hu (detection vs load)
posterior_summary(m1, variable = "^cor_replicate_group", regex = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### Detection and noise
Left: posterior median of each replicate group's latent load against its detection probability. Right: noise SD by noise type (posterior median, 50% and 95% intervals).
```{r m1-curves}
#| code-fold: true
#| code-summary: "Show Detection and noise code"
#| fig-width: 9
#| fig-height: 4

m1_groups = m1_df %>%
  distinct(replicate_group, .keep_all = TRUE)

m1_group_est = m1_groups %>%
  mutate(
    log10_load = apply(posterior_linpred(m1, newdata = m1_groups, dpar = "mu"), 2, median) / log(10),
    p_detect = 1 - apply(posterior_linpred(m1, newdata = m1_groups, dpar = "hu",
                                           transform = TRUE), 2, median))

p_det = ggplot(m1_group_est, aes(log10_load, p_detect, colour = qpcr_lab)) +
  geom_point(alpha = 0.6) +
  labs(x = "latent log10(Bd load)", y = "P(detected)", colour = "qPCR lab") +
  theme_bw() +
  theme(legend.position = "bottom")

# sigma is on the log scale of ln(load); convert to SD of log10(load)
m1_sigma = as_draws_df(m1) %>%
  as_tibble() %>%
  select(starts_with("b_sigma_noise_type")) %>%
  pivot_longer(everything(), names_to = "noise_type", values_to = "value") %>%
  mutate(noise_type = str_remove(noise_type, "b_sigma_noise_type"),
         sigma_log10 = exp(value) / log(10)) %>%
  group_by(noise_type) %>%
  summarise(med = median(sigma_log10),
            lo50 = quantile(sigma_log10, 0.25), hi50 = quantile(sigma_log10, 0.75),
            lo95 = quantile(sigma_log10, 0.025), hi95 = quantile(sigma_log10, 0.975),
            .groups = "drop")

p_sig = ggplot(m1_sigma, aes(noise_type, med)) +
  geom_linerange(aes(ymin = lo95, ymax = hi95)) +
  geom_linerange(aes(ymin = lo50, ymax = hi50), linewidth = 1.5) +
  geom_point(size = 2.5) +
  labs(x = "noise type", y = "noise SD (log10)") +
  theme_bw()

p_det + p_sig
```

### Posterior predictive checks
Share of results detected, and median positive log10 load, by replicate type × qPCR lab.
```{r m1-ppc}
#| code-fold: true
#| code-summary: "Show PPC code"
#| fig-width: 8
#| fig-height: 4

# pp_check only sees model variables, so call bayesplot directly with the grouping column
m1_yrep = posterior_predict(m1, ndraws = 500)

ppc_stat_grouped(y = m1_df$bd_load, yrep = m1_yrep, group = m1_df$ppc_group,
                 stat = function(y) mean(y > 0)) +
  labs(x = "share detected")

ppc_stat_grouped(y = m1_df$bd_load, yrep = m1_yrep, group = m1_df$ppc_group,
                 stat = function(y) median(log10(y[y > 0]))) +
  labs(x = "median log10(Bd load) among positives")
```

### Export for Model 2
Posterior draws and summaries (mean, SD, median, 95% interval) of population-level effects, group-level SDs, and correlations, saved to `fits/m1_replicates_export.rds` for use as priors in Model 2.
```{r m1-export}
#| code-fold: true
#| code-summary: "Show Export code"

m1_draws = as_draws_df(m1) %>%
  subset_draws(variable = c("^b_", "^sd_", "^cor_"), regex = TRUE)

m1_summary = m1_draws %>%
  summarise_draws(mean, sd, median,
                  ~ quantile(.x, probs = c(0.025, 0.975)))

m1_export = list(summary = m1_summary,
                 draws = m1_draws,
                 formula = m1_formula,
                 import_date = import_date,
                 export_date = Sys.Date())

saveRDS(m1_export, here(fit_dir, "m1_replicates_export.rds"))

m1_summary %>%
  kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```
