---
title: "rz_snarl_qpcr_model2_experiment"
author: "Cob Staines"
format: html
editor: visual
---

# RZ-SNARL qPCR Comparison: Model 2 (Experiment)

**Objective**: Quantify between-lab differences in extraction and qPCR from the crossed experiment, with priors on noise and detection informed by Model 1 (`rz_snarl_qpcr_model1_replicates.qmd`).

## Setup

Load packages; load and clean experiment and replicate data (`R/data_prep.R`, from data saved by `R/data_import.R`)

```{r setup}
#| code-fold: true
#| code-summary: "Show Setup code"
#| results: hide
librarian::shelf(tidyverse, here, knitr, kableExtra, ggplot2, patchwork, brms, cmdstanr, posterior, bayesplot)

cmdstanr::check_cmdstan_toolchain()   # confirms Stan is found
set_cmdstan_path(Sys.getenv("cmd_stan_path"))

source(here("R", "data_prep.R"))

# fitted models are saved outside the repo
fit_dir = here(wddir, "fits")
dir.create(fit_dir, showWarnings = FALSE)

# set TRUE to refit models instead of loading saved fits
refit = FALSE
```

## Model 1 results
Posterior summaries exported by Model 1 (`fits/m1_replicates_export.rds`); render Model 1 first.
```{r m1-import}
#| code-fold: true
#| code-summary: "Show Model 1 import code"

m1_export = readRDS(here(fit_dir, "m1_replicates_export.rds"))
m1_summary = m1_export$summary
m1_draws = m1_export$draws

if (m1_export$import_date != import_date) {
  warning("Model 1 was fit on data imported ", m1_export$import_date,
          "; current data imported ", import_date, ". Consider re-rendering Model 1.")
}

m1_summary %>%
  kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## Model 2: Experiment
To be implemented (brms): latent true load per frog (group-level intercept correlated across `mu` and `hu`); lab-specific extraction and qPCR effects on `mu` and `hu`, with priors on noise and detection informed by Model 1.
