---
title: "rz_snarl_qpcr_model2_experiment"
author: "Cob Staines"
format: html
editor: visual
---

# RZ-SNARL qPCR Comparison: Model 2 (Experiment)

**Objective**: Quantify between-lab differences in extraction and qPCR from the crossed experiment, with priors on noise and detection informed by Model 1 (`rz_snarl_qpcr_model1_replicates.qmd`).

## Setup

Load packages; load and clean experiment and replicate data (`R/data_prep.R`, from data saved by `R/data_import.R`)

```{r setup}
#| code-fold: true
#| code-summary: "Show Setup code"
#| results: hide
librarian::shelf(tidyverse, here, knitr, kableExtra, ggplot2, patchwork, brms, cmdstanr, posterior, bayesplot)

cmdstanr::check_cmdstan_toolchain()   # confirms Stan is found
set_cmdstan_path(Sys.getenv("cmd_stan_path"))

source(here("R", "data_prep.R"))

# fitted models are saved outside the repo
fit_dir = here(wddir, "fits")
dir.create(fit_dir, showWarnings = FALSE)

# set TRUE to refit models instead of loading saved fits
refit = FALSE
```

## Model 1 results
Posterior summaries exported by Model 1 (`fits/m1_replicates_export.rds`); render Model 1 first.
```{r m1-import}
#| code-fold: true
#| code-summary: "Show Model 1 import code"

m1_export = readRDS(here(fit_dir, "m1_replicates_export.rds"))
m1_summary = m1_export$summary
m1_draws = m1_export$draws

if (m1_export$import_date != import_date) {
  warning("Model 1 was fit on data imported ", m1_export$import_date,
          "; current data imported ", import_date, ". Consider re-rendering Model 1.")
}

m1_summary %>%
  kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## Model 2: Experiment
Between-lab extraction and qPCR differences from the crossed experiment, as a brms non-linear hurdle-lognormal model (cmdstanr backend). All quantities are on the natural-log scale of Bd load (ITS1 copies per swab) unless noted; results are converted to log10 for reporting.

For positive results, the expected log load is

$$\mu = L + d + s \cdot \text{SNARL qPCR} \cdot (L - L_\text{ref})$$

- **`L`, latent swab load**: an intercept, a frog-level intercept (the frog's true load), and a swab-level intercept (swab + extraction noise; shared by results 1 & 2, which share an extract), with an SD pooled across extraction methods. A version with a separate SD per extraction method showed the two SDs strongly negatively correlated in the posterior: each frog has exactly one Qiagen and one PrepMan swab, so the design identifies only their combined variance. Whether one extraction method is noisier than the other cannot be answered with this design.
- **`d`, constant biases**: PrepMan extraction (vs Qiagen) and SNARL qPCR (vs RZ). The reference pipeline is Qiagen extraction with RZ qPCR.
- **`s`, SNARL qPCR slope deviation**: SNARL's reading departs from RZ's by `d_snarl + s·(L − L_ref)`. `s > 0` means SNARL reads increasingly low as load drops below `L_ref` (the hypothesis: SNARL underestimates low loads, otherwise comparable); `s = 0` means a constant bias. `L_ref` is the median positive load in the experiment, so `d_snarl` is the bias at a typical load.
- **`hu`, detection**: intercepts by extraction method and qPCR lab, with a frog-level intercept correlated with the frog-level intercept in `L` (shared ID `g`), as in Model 1. Load-dependent detection enters through this correlation, so labs differ in baseline detection but share the detection-vs-load slope.
- **`sigma`, residual (qPCR) noise**, pooled across qPCR labs. Extract-level noise is in the swab effect, so the residual is qPCR noise. A first version estimated `sigma` by qPCR lab (SNARL ≈ 0.38, RZ ≈ 0.16 log10), but the design mainly identifies the *sum* of the two labs' qPCR noise (results 1 & 2 share an extract); separating them relied on single-result swabs, where qPCR noise is confounded with swab noise. RZ `sigma` had low ESS (205) and the fit had 5 divergent transitions, so lab-specific qPCR noise is treated as exploratory only.

Not included: population (`collect_lab`) effects in `L` and `hu`. The focus is on extraction and qPCR lab differences, which are estimated within frogs; population differences in load and detection (Sierra frogs ≈ 2.4 log10 higher loads in an earlier version) are absorbed by the frog-level intercepts. An earlier version with `collect_lab` estimated PrepMan −0.38 log10 and SNARL qPCR +0.06 log10; these should change little.

Not included: an extraction × qPCR interaction (e.g. PrepMan extracts behaving differently in RZ qPCR). All four pipeline combinations are observed, but each pair of combinations compares different frog populations (see README panel table), so an interaction would be confounded with population; left out of this first version.

### Model data
```{r m2-data}
#| code-fold: true
#| code-summary: "Show Model data code"

m2_df = snarl_clean %>%
  filter(!is.na(bd_load)) %>%
  mutate(collect_lab = factor(collect_lab, levels = c("rz", "snarl")),
         extraction_method = factor(extraction_method, levels = c("qiagen", "prepman")),
         qpcr_lab = factor(qpcr_lab, levels = c("rz", "snarl")),
         is_prepman = as.integer(extraction_method == "prepman"),
         is_snarl_qpcr = as.integer(qpcr_lab == "snarl"),
         pipeline = paste(extraction_method, qpcr_lab, sep = "_"))

# reference load (ln scale): median positive load
L_ref = median(log(m2_df$bd_load[m2_df$bd_load > 0]))
m2_df = m2_df %>%
  mutate(L_ref = L_ref)

m2_df %>%
  group_by(extraction_method, qpcr_lab) %>%
  summarise(frogs = n_distinct(frog_id),
            results = n(),
            detected = mean(bd_load > 0),
            median_log10_pos = median(log10(bd_load[bd_load > 0])),
            .groups = "drop") %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

`L_ref` = `r round(L_ref / log(10), 2)` (log10 copies per swab).

### Priors from Model 1
Informative priors only where Model 1 is well identified and the parameter means the same thing in both models (SDs widened ×2):

- **Residual (qPCR) noise** `sigma`: Model 1 qPCR replicate noise.
- **Swab SD** in `L`: Model 1 swab replicate noise minus qPCR noise, $\sqrt{\sigma_\text{swab}^2 - \sigma_\text{qPCR}^2}$.
- **Frog-level `hu` SD**: Model 1 group-level `hu` SD.

Detection and load intercepts get weakly informative priors (they depend on each dataset's load distribution). The `L`–`hu` correlation gets `lkj(2)`: brms only supports LKJ priors on correlations, so Model 1's correlation (≈ −0.55) cannot be used as a centred prior.
```{r m2-priors-from-m1}
#| code-fold: true
#| code-summary: "Show Priors from Model 1 code"

m1_mean_sd = function(var) {
  row = filter(m1_summary, variable == var)
  c(mean = row$mean, sd = row$sd)
}

sigma_qpcr_m1 = m1_mean_sd("b_sigma_noise_typeqpcr")
sd_hu_m1 = m1_mean_sd("sd_replicate_group__hu_Intercept")

# swab + extraction SD (ln scale) = sqrt(total swab noise^2 - qPCR noise^2), per draw
swab_sd_draws = sqrt(pmax(exp(2 * m1_draws$b_sigma_noise_typeswab) -
                            exp(2 * m1_draws$b_sigma_noise_typeqpcr), 0))
swab_sd_m1 = c(mean = mean(swab_sd_draws), sd = sd(swab_sd_draws))

tibble(prior = c("sigma (log scale)", "sd swab (L)", "sd frog (hu)"),
       m1_mean = c(sigma_qpcr_m1["mean"], swab_sd_m1["mean"], sd_hu_m1["mean"]),
       m1_sd = c(sigma_qpcr_m1["sd"], swab_sd_m1["sd"], sd_hu_m1["sd"])) %>%
  mutate(prior_sd = 2 * m1_sd) %>%
  kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### Model structure
```{r m2-structure}
#| code-fold: true
#| code-summary: "Show Model structure code"

m2_family = hurdle_lognormal()

m2_formula = bf(
  bd_load ~ L + d + s * is_snarl_qpcr * (L - L_ref),

  L ~ 1 +
    (1 | g | frog_id) +
    (1 | bd_swab_id),

  d ~ 0 + is_prepman + is_snarl_qpcr,

  s ~ 1,

  hu ~ 1 + extraction_method + qpcr_lab +
    (1 | g | frog_id),

  sigma ~ 1,

  nl = TRUE
)

m2_priors = c(
  # L (ln copies per swab)
  prior(normal(7, 4), class = "b", nlpar = "L", coef = "Intercept"),
  prior(normal(0, 3), class = "sd", nlpar = "L", group = "frog_id"),
  prior_string(paste0("normal(", swab_sd_m1["mean"], ", ", 2 * swab_sd_m1["sd"], ")"),
               class = "sd", nlpar = "L", group = "bd_swab_id"),

  # d: constant biases (ln scale)
  prior(normal(0, 2), class = "b", nlpar = "d"),

  # s: SNARL qPCR slope deviation
  prior(normal(0, 0.5), class = "b", nlpar = "s"),

  # hu
  prior(normal(0, 2), class = "Intercept", dpar = "hu"),
  prior(normal(0, 2), class = "b", dpar = "hu"),
  prior_string(paste0("normal(", sd_hu_m1["mean"], ", ", 2 * sd_hu_m1["sd"], ")"),
               class = "sd", dpar = "hu", group = "frog_id"),

  # sigma (log of ln-scale SD), from Model 1 qPCR noise
  prior_string(paste0("normal(", sigma_qpcr_m1["mean"], ", ", 2 * sigma_qpcr_m1["sd"], ")"),
               class = "Intercept", dpar = "sigma"),

  # correlations
  prior(lkj(2), class = "cor")
)
```

### Fit
Fit is saved to `fits/` in the data directory and reloaded unless `refit = TRUE`.
```{r m2-fit}
#| code-fold: true
#| code-summary: "Show Fit code"
#| results: hide

m2 = brm(
  formula = m2_formula,
  data = m2_df,
  family = m2_family,
  prior = m2_priors,
  control = list(adapt_delta = 0.99),
  chains = 4,
  iter = 4000,
  warmup = 2000,
  seed = 1234,
  cores = 4,
  backend = "cmdstanr",
  file = here(fit_dir, "m2_experiment"),
  file_refit = if (refit) "always" else "on_change"
)
```

### Convergence & summary
```{r m2-summary}
#| code-fold: true
#| code-summary: "Show Summary code"

summary(m2)
```

### Divergences
Pairs plot of noise, bias, and slope parameters, with divergent transitions highlighted (red), to locate any remaining sampling problems.
```{r m2-divergences}
#| code-fold: true
#| code-summary: "Show Divergences code"
#| fig-width: 9
#| fig-height: 9

m2_np = nuts_params(m2)
cat("Divergent transitions:", sum(m2_np$Value[m2_np$Parameter == "divergent__"]), "\n")

mcmc_pairs(as.array(m2), np = m2_np,
           pars = c("b_sigma_Intercept",
                    "sd_bd_swab_id__L_Intercept",
                    "sd_frog_id__L_Intercept",
                    "b_d_is_snarl_qpcr",
                    "b_s_Intercept"),
           off_diag_args = list(size = 0.5, alpha = 0.3))
```

### Lab differences
Key effects, converted to log10 where on the load scale (posterior median, 95% interval, and posterior probability of the effect being positive).

- `d_prepman`: PrepMan vs Qiagen extraction bias (log10).
- `d_snarl`: SNARL vs RZ qPCR bias at `L_ref` (log10).
- `s`: SNARL qPCR slope deviation (unitless; `> 0` = SNARL reads increasingly low below `L_ref`).
- `hu_*`: detection differences (logit scale of P(zero); `> 0` = more zeros).
- `sigma`: residual qPCR noise SD, pooled across labs (log10).
- `sd_swab`: swab + extraction noise SD, pooled across extraction methods (log10).
```{r m2-effects}
#| code-fold: true
#| code-summary: "Show Lab differences code"

m2_draws = as_draws_df(m2) %>%
  as_tibble()

m2_effects = m2_draws %>%
  transmute(d_prepman = b_d_is_prepman / log(10),
            d_snarl = b_d_is_snarl_qpcr / log(10),
            s = b_s_Intercept,
            hu_prepman = b_hu_extraction_methodprepman,
            hu_snarl_qpcr = b_hu_qpcr_labsnarl,
            sigma = exp(b_sigma_Intercept) / log(10),
            sd_swab = sd_bd_swab_id__L_Intercept / log(10),
            cor_L_hu = cor_frog_id__hu_Intercept__L_Intercept)

m2_effects %>%
  pivot_longer(everything(), names_to = "effect") %>%
  group_by(effect) %>%
  summarise(median = median(value),
            lo95 = quantile(value, 0.025),
            hi95 = quantile(value, 0.975),
            p_positive = mean(value > 0),
            .groups = "drop") %>%
  kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### SNARL vs RZ qPCR
Expected difference (SNARL − RZ, log10) in the reading of the same extract, across latent load: `d_snarl + s·(L − L_ref)`. A flat line at 0 means the labs agree; a line rising with load means SNARL underestimates low loads. Right: detection probability by pipeline at the average frog.
```{r m2-snarl-rz}
#| code-fold: true
#| code-summary: "Show SNARL vs RZ code"
#| fig-width: 9
#| fig-height: 4

load_grid = seq(1, 7, by = 0.1)  # log10

qpcr_diff = expand_grid(m2_draws %>% select(.draw, b_d_is_snarl_qpcr, b_s_Intercept),
                        log10_load = load_grid) %>%
  mutate(diff_log10 = (b_d_is_snarl_qpcr + b_s_Intercept * (log10_load * log(10) - L_ref)) / log(10)) %>%
  group_by(log10_load) %>%
  summarise(med = median(diff_log10),
            lo = quantile(diff_log10, 0.05),
            hi = quantile(diff_log10, 0.95),
            .groups = "drop")

p_mu = ggplot(qpcr_diff, aes(log10_load, med)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = L_ref / log(10), linetype = "dotted", colour = "grey50") +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.4) +
  geom_line() +
  labs(x = "latent log10(Bd load)", y = "SNARL − RZ qPCR (log10)") +
  theme_bw()

# detection probability at the average frog, by pipeline
detect_df = m2_draws %>%
  transmute(.draw,
            qiagen_rz = 1 - plogis(b_hu_Intercept),
            qiagen_snarl = 1 - plogis(b_hu_Intercept + b_hu_qpcr_labsnarl),
            prepman_rz = 1 - plogis(b_hu_Intercept + b_hu_extraction_methodprepman),
            prepman_snarl = 1 - plogis(b_hu_Intercept + b_hu_extraction_methodprepman + b_hu_qpcr_labsnarl)) %>%
  pivot_longer(-.draw, names_to = "pipeline", values_to = "p_detect") %>%
  group_by(pipeline) %>%
  summarise(med = median(p_detect),
            lo = quantile(p_detect, 0.05),
            hi = quantile(p_detect, 0.95),
            .groups = "drop")

p_hu = ggplot(detect_df, aes(pipeline, med)) +
  geom_linerange(aes(ymin = lo, ymax = hi)) +
  geom_point(size = 2.5) +
  labs(x = "extraction – qPCR lab", y = "P(detected), average frog") +
  theme_bw()

p_mu + p_hu
```

### Posterior predictive checks
Share of results detected, and median positive log10 load, by pipeline (extraction method × qPCR lab).
```{r m2-ppc}
#| code-fold: true
#| code-summary: "Show PPC code"
#| fig-width: 8
#| fig-height: 4

m2_yrep = posterior_predict(m2, ndraws = 500)

ppc_stat_grouped(y = m2_df$bd_load, yrep = m2_yrep, group = m2_df$pipeline,
                 stat = function(y) mean(y > 0)) +
  labs(x = "share detected")

ppc_stat_grouped(y = m2_df$bd_load, yrep = m2_yrep, group = m2_df$pipeline,
                 stat = function(y) median(log10(y[y > 0]))) +
  labs(x = "median log10(Bd load) among positives")
```

### Model comparison: slope vs constant bias
Model 2b is Model 2 without the SNARL qPCR slope `s` (constant biases only: $\mu = L + d$), otherwise identical (same data, priors, and settings). The models are compared by approximate leave-one-out cross-validation (PSIS-LOO). If the difference in expected log predictive density (`elpd_diff`) is small relative to its SE (roughly < 2 SE), the simpler constant-bias model is preferred.
```{r m2b-fit}
#| code-fold: true
#| code-summary: "Show Model 2b fit code"
#| results: hide

m2b_formula = bf(
  bd_load ~ L + d,

  L ~ 1 +
    (1 | g | frog_id) +
    (1 | bd_swab_id),

  d ~ 0 + is_prepman + is_snarl_qpcr,

  hu ~ 1 + extraction_method + qpcr_lab +
    (1 | g | frog_id),

  sigma ~ 1,

  nl = TRUE
)

# same priors, minus the slope
m2b_priors = m2_priors[m2_priors$nlpar != "s", ]

m2b = brm(
  formula = m2b_formula,
  data = m2_df,
  family = m2_family,
  prior = m2b_priors,
  control = list(adapt_delta = 0.99),
  chains = 4,
  iter = 4000,
  warmup = 2000,
  seed = 1234,
  cores = 4,
  backend = "cmdstanr",
  file = here(fit_dir, "m2b_experiment_constant"),
  file_refit = if (refit) "always" else "on_change"
)

# loo is stored in the saved fit files, so it is only computed once
m2 = add_criterion(m2, "loo")
m2b = add_criterion(m2b, "loo")
```

```{r m2-compare}
#| code-fold: true
#| code-summary: "Show Model comparison code"

summary(m2b)

# Pareto k diagnostics: values > 0.7 mean the LOO estimate for that result is unreliable
tibble(model = c("m2 (slope)", "m2b (constant)"),
       n_k_over_0.7 = c(sum(loo(m2)$diagnostics$pareto_k > 0.7),
                        sum(loo(m2b)$diagnostics$pareto_k > 0.7))) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

loo_compare(m2, m2b) %>%
  as.data.frame() %>%
  rownames_to_column("model") %>%
  select(model, elpd_diff, se_diff, elpd_loo, p_loo) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

# constant correction to the Qiagen-RZ scale under Model 2b (log10 and multiplier)
m2b_draws = as_draws_df(m2b) %>%
  as_tibble()

m2b_correction = m2b_draws %>%
  transmute(prepman_snarl = -(b_d_is_prepman + b_d_is_snarl_qpcr) / log(10),
            prepman_rz = -b_d_is_prepman / log(10),
            qiagen_snarl = -b_d_is_snarl_qpcr / log(10))

m2b_correction %>%
  pivot_longer(everything(), names_to = "pipeline", values_to = "correction_log10") %>%
  group_by(pipeline) %>%
  summarise(median = median(correction_log10),
            lo95 = quantile(correction_log10, 0.025),
            hi95 = quantile(correction_log10, 0.975),
            multiplier = 10^median,
            multiplier_lo95 = 10^lo95,
            multiplier_hi95 = 10^hi95,
            .groups = "drop") %>%
  kable(digits = 2, caption = "Model 2b: constant correction to Qiagen–RZ scale (add to log10 load, or multiply copies)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

# export constant correction draws (log10) for applying elsewhere
saveRDS(list(draws = m2b_correction,
             reference_pipeline = "qiagen_rz",
             scale = "log10 correction to add to positive log10 loads",
             import_date = import_date,
             export_date = Sys.Date()),
        here(fit_dir, "m2b_constant_correction_export.rds"))
```

### Detection conversion (Model 2b)
Probability of detecting Bd by pipeline, as a function of a frog's true load on the reference (Qiagen–RZ) scale. Derived from the frog-level `L`–`hu` correlation: for a frog with true load $\ell$ (ln),

$$\text{hu} \mid \ell \sim N\left(h_0 + h_\text{pipeline} + \rho \frac{\sigma_{hu}}{\sigma_L}(\ell - L_0),\; \sigma_{hu}^2(1 - \rho^2)\right)$$

averaged over the conditional spread (probit approximation to the logistic-normal), $P(\text{detected}) = 1 - E[\text{logit}^{-1}(\text{hu})]$.

Caveats: pipelines share the detection-vs-load slope and differ only in baseline (correlation approach); the curve reflects frog-level load (swab-to-swab variation not included); population differences are absorbed by the frog-level effects, so the load–detection relationship mixes within- and between-population variation; curves are extrapolated below the range of observed positive loads.
```{r m2b-detection}
#| code-fold: true
#| code-summary: "Show Detection conversion code"
#| fig-width: 7
#| fig-height: 4.5

# thin to 2000 draws to keep the grid manageable
m2b_det_draws = m2b_draws %>%
  slice(round(seq(1, n(), length.out = 2000))) %>%
  transmute(.draw,
            L_0 = b_L_Intercept,
            hu_0 = b_hu_Intercept,
            hu_prepman = b_hu_extraction_methodprepman,
            hu_snarl = b_hu_qpcr_labsnarl,
            sd_L = sd_frog_id__L_Intercept,
            sd_hu = sd_frog_id__hu_Intercept,
            rho = cor_frog_id__hu_Intercept__L_Intercept)

det_pipelines = tribble(
  ~pipeline,        ~prepman, ~snarl_qpcr,
  "qiagen_rz",      0,        0,
  "qiagen_snarl",   0,        1,
  "prepman_rz",     1,        0,
  "prepman_snarl",  1,        1)

detection_draws = expand_grid(m2b_det_draws,
                              det_pipelines,
                              log10_load = seq(0, 9, by = 0.05)) %>%
  mutate(hu_mean = hu_0 + hu_prepman * prepman + hu_snarl * snarl_qpcr +
           rho * sd_hu / sd_L * (log10_load * log(10) - L_0),
         hu_var = sd_hu^2 * (1 - rho^2),
         p_detect = 1 - plogis(hu_mean / sqrt(1 + pi * hu_var / 8))) %>%
  select(.draw, pipeline, log10_load, p_detect)

detection_curves = detection_draws %>%
  group_by(pipeline, log10_load) %>%
  summarise(med = median(p_detect),
            lo = quantile(p_detect, 0.05),
            hi = quantile(p_detect, 0.95),
            .groups = "drop")

# range of observed positive loads in the experiment (log10), for reference
obs_range = range(log10(m2_df$bd_load[m2_df$bd_load > 0]))

detection_curves %>%
  filter(log10_load <= 7) %>%
  ggplot(aes(log10_load, med, colour = pipeline, fill = pipeline)) +
  annotate("rect", xmin = obs_range[1], xmax = min(obs_range[2], 7), ymin = -Inf, ymax = Inf,
           fill = "grey90", alpha = 0.5) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line() +
  labs(x = "true log10(Bd load), Qiagen–RZ scale",
       y = "P(detected)",
       colour = NULL, fill = NULL,
       caption = "Shaded: range of observed positive loads in the experiment") +
  theme_bw() +
  theme(legend.position = "bottom")
```

Detection thresholds: true load (log10, Qiagen–RZ scale) at which each pipeline reaches 50% and 90% detection (posterior median and 95% interval).
```{r m2b-detection-thresholds}
#| code-fold: true
#| code-summary: "Show Detection thresholds code"

detection_thresholds = detection_draws %>%
  group_by(.draw, pipeline) %>%
  summarise(load_50 = suppressWarnings(min(log10_load[p_detect >= 0.5])),
            load_90 = suppressWarnings(min(log10_load[p_detect >= 0.9])),
            .groups = "drop") %>%
  pivot_longer(c(load_50, load_90), names_to = "threshold", values_to = "log10_load")

detection_thresholds %>%
  group_by(pipeline, threshold) %>%
  summarise(median = median(log10_load[is.finite(log10_load)]),
            lo95 = quantile(log10_load[is.finite(log10_load)], 0.025),
            hi95 = quantile(log10_load[is.finite(log10_load)], 0.975),
            # draws never reaching the threshold within log10 load 0-9
            share_not_reached = mean(!is.finite(log10_load)),
            .groups = "drop") %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

Detection probability at selected true loads for the two labs' standard pipelines, and the difference (RZ − SNARL; posterior median and 95% interval).
```{r m2b-detection-table}
#| code-fold: true
#| code-summary: "Show Detection table code"

detection_draws %>%
  filter(pipeline %in% c("qiagen_rz", "prepman_snarl"),
         round(log10_load, 2) %in% 1:6) %>%
  pivot_wider(names_from = pipeline, values_from = p_detect) %>%
  mutate(difference = qiagen_rz - prepman_snarl) %>%
  pivot_longer(c(qiagen_rz, prepman_snarl, difference), names_to = "quantity") %>%
  group_by(log10_load, quantity) %>%
  summarise(value = sprintf("%.2f (%.2f, %.2f)", median(value),
                            quantile(value, 0.025), quantile(value, 0.975)),
            .groups = "drop") %>%
  pivot_wider(names_from = quantity, values_from = value) %>%
  select(log10_load, qiagen_rz, prepman_snarl, difference) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

# export detection curves and thresholds
saveRDS(list(curves = detection_curves,
             thresholds = detection_thresholds,
             load_scale = "true log10 load, Qiagen-RZ scale (frog level)",
             import_date = import_date,
             export_date = Sys.Date()),
        here(fit_dir, "m2b_detection_export.rds"))
```

### Correction to a common scale (Model 2, with slope)
Conversion of a positive reading from any pipeline to the reference pipeline (Qiagen extraction, RZ qPCR). For a reading $y$ (ln scale) from a pipeline with extraction bias $d_\text{ext}$ (PrepMan only), qPCR bias $d_\text{q}$ and slope $s$ (SNARL qPCR only):

$$y = L + d_\text{ext} + d_\text{q} + s(L - L_\text{ref}) \quad\Rightarrow\quad L = \frac{y - d_\text{ext} - d_\text{q} + s L_\text{ref}}{1 + s}$$

where $L$ is the reference-pipeline (Qiagen + RZ) equivalent. The plot and table show the correction $L - y$ (log10) across observed loads (posterior median, 90% interval). Corrections apply to positive results only: zeros cannot be corrected, and pipelines differ in detection (see detection plot above).
```{r m2-correction}
#| code-fold: true
#| code-summary: "Show Correction code"
#| fig-width: 7
#| fig-height: 4

# posterior draws of correction parameters (ln scale)
m2_correction_draws = m2_draws %>%
  transmute(.draw,
            d_prepman = b_d_is_prepman,
            d_snarl = b_d_is_snarl_qpcr,
            s = b_s_Intercept)

pipelines = tribble(
  ~pipeline,        ~prepman, ~snarl_qpcr,
  "prepman_snarl",  1,        1,
  "prepman_rz",     1,        0,
  "qiagen_snarl",   0,        1)

correction = expand_grid(m2_correction_draws,
                         pipelines,
                         log10_obs = seq(1, 7, by = 0.1)) %>%
  mutate(y = log10_obs * log(10),
         slope = s * snarl_qpcr,
         L = (y - d_prepman * prepman - d_snarl * snarl_qpcr + slope * L_ref) / (1 + slope),
         correction_log10 = (L - y) / log(10)) %>%
  group_by(pipeline, log10_obs) %>%
  summarise(med = median(correction_log10),
            lo = quantile(correction_log10, 0.05),
            hi = quantile(correction_log10, 0.95),
            .groups = "drop")

ggplot(correction, aes(log10_obs, med, colour = pipeline, fill = pipeline)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line() +
  labs(x = "observed log10(Bd load)",
       y = "correction to Qiagen–RZ scale (log10)",
       colour = NULL, fill = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")

correction %>%
  filter(log10_obs %in% c(2, 3, 4, 5, 6)) %>%
  mutate(correction = sprintf("%.2f (%.2f, %.2f)", med, lo, hi)) %>%
  select(pipeline, log10_obs, correction) %>%
  pivot_wider(names_from = log10_obs, values_from = correction,
              names_prefix = "log10 load = ") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

# export correction parameter draws (ln scale) and L_ref for applying the correction elsewhere
saveRDS(list(draws = m2_correction_draws,
             L_ref = L_ref,
             reference_pipeline = "qiagen_rz",
             import_date = import_date,
             export_date = Sys.Date()),
        here(fit_dir, "m2_correction_export.rds"))
```
