---
title: "rz_snarl_qpcr_model1_replicates"
author: "Cob Staines"
format: html
editor: visual
---

# RZ-SNARL qPCR Comparison: Model 1 (Replicates)

**Objective**: Model within-lab noise and detection from replicates. Posterior draws and summaries are exported for use as priors in Model 2 (`rz_snarl_qpcr_model2_experiment.qmd`).

## Setup

Load packages; load and clean experiment and replicate data (`R/data_prep.R`, from data saved by `R/data_import.R`)

```{r setup}
#| code-fold: true
#| code-summary: "Show Setup code"
#| results: hide
librarian::shelf(tidyverse, here, knitr, kableExtra, ggplot2, patchwork, brms, cmdstanr, posterior, bayesplot)

cmdstanr::check_cmdstan_toolchain()   # confirms Stan is found
set_cmdstan_path(Sys.getenv("cmd_stan_path"))

source(here("R", "data_prep.R"))

# fitted models are saved outside the repo
fit_dir = here(wddir, "fits")
dir.create(fit_dir, showWarnings = FALSE)

# set TRUE to refit models instead of loading saved fits
refit = FALSE
```

## Model 1: Replicates

Within-lab noise and detection as a function of load, as a brms hurdle-lognormal model (cmdstanr backend).

- **Latent load**: each replicate group (a capture for swab replicates, an extract for qPCR replicates) has a group-level intercept in `mu`, i.e. its true (log) load.
- **Detection**: `hu` (probability of a zero) has a qPCR lab effect and a group-level intercept. The group-level intercepts of `mu` and `hu` are correlated (shared ID `g`), so the `mu`–`hu` correlation captures load-dependent detection (expected negative: higher load, fewer zeros).
- **Noise**: `sigma` differs by noise type: swab replicates (total noise: swab + extraction + qPCR) and qPCR replicates (total noise between qPCR re-runs of the same extract).

Kept deliberately simple given limited replicates. An earlier version also had a qPCR run (extract × plate) effect and a `sigma` group-level intercept (load-dependent noise); neither was identified (nearly every qPCR run was a single well; 2–3 positives per group), and they caused divergent transitions.

`mu` and `sigma` are on the natural-log scale of Bd load (ITS1 copies per swab); results are converted to log10 for plots.

### Model data

```{r m1-data}
#| code-fold: true
#| code-summary: "Show Model data code"

m1_df = replicates_clean %>%
  mutate(qpcr_lab = factor(qpcr_lab, levels = c("rz", "snarl")),
         noise_type = factor(replicate_type, levels = c("swab", "qpcr")),
         ppc_group = paste(replicate_type, qpcr_lab))

m1_df %>%
  group_by(replicate_type, qpcr_lab) %>%
  summarise(groups = n_distinct(replicate_group),
            results = n(),
            qpcr_plates = n_distinct(qpcr_plate_name),
            detected = mean(bd_load > 0),
            .groups = "drop") %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### Model structure

```{r m1-structure}
#| code-fold: true
#| code-summary: "Show Model structure code"

m1_family = hurdle_lognormal()

m1_formula = bf(
  bd_load ~ 1 +
    (1 | g | replicate_group),

  hu ~ 1 + qpcr_lab +
    (1 | g | replicate_group),

  sigma ~ 0 + noise_type
)

m1_priors = c(
  # mu (ln copies per swab)
  prior(normal(7, 4), class = "Intercept"),

  # hu
  prior(normal(0, 2), class = "Intercept", dpar = "hu"),
  prior(normal(0, 2), class = "b", dpar = "hu"),

  # sigma (log of ln-scale SD)
  prior(normal(0, 1), class = "b", dpar = "sigma"),

  # group-level effects
  prior(exponential(1), class = "sd"),
  prior(lkj(2), class = "cor")
)
```

### Fit

Fit is saved to `fits/` in the data directory and reloaded unless `refit = TRUE`.

```{r m1-fit}
#| code-fold: true
#| code-summary: "Show Fit code"
#| results: hide

m1 = brm(
  formula = m1_formula,
  data = m1_df,
  family = m1_family,
  prior = m1_priors,
  control = list(adapt_delta = 0.98),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 1234,
  cores = 4,
  backend = "cmdstanr",
  file = here(fit_dir, "m1_replicates"),
  file_refit = if (refit) "always" else "on_change"
)
```

### Convergence & summary

```{r m1-summary}
#| code-fold: true
#| code-summary: "Show Summary code"

summary(m1)

# group-level correlation: mu-hu (detection vs load)
posterior_summary(m1, variable = "^cor_replicate_group", regex = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### Detection and noise

Left: each replicate group's latent load (posterior median, 90% interval) against its detection probability (posterior median), coloured by the share of its results detected. Groups with the same detection outcome fall on the same strand, since with 2–3 results per group the detected fraction dominates each group's `hu` estimate. The line and band are the population-level detection curve implied by the model, $P(\text{detected} \mid \text{load})$, for RZ qPCR (posterior median, 90% interval). Right: noise SD by noise type (posterior median, 50% and 95% intervals).

```{r m1-curves}
#| code-fold: true
#| code-summary: "Show Detection and noise code"
#| fig-width: 10
#| fig-height: 4.5

m1_groups = m1_df %>%
  group_by(replicate_group) %>%
  mutate(detected_share = paste0(sum(bd_load > 0), "/", n())) %>%
  slice_head(n = 1) %>%
  ungroup()

mu_group = posterior_linpred(m1, newdata = m1_groups, dpar = "mu") / log(10)
hu_group = posterior_linpred(m1, newdata = m1_groups, dpar = "hu", transform = TRUE)

m1_group_est = m1_groups %>%
  mutate(log10_load = apply(mu_group, 2, median),
         log10_load_lo = apply(mu_group, 2, quantile, 0.05),
         log10_load_hi = apply(mu_group, 2, quantile, 0.95),
         p_detect = 1 - apply(hu_group, 2, median))

# population-level detection curve (RZ qPCR), from the bivariate normal group effects:
# hu | mu ~ N(hu_0 + rho * sd_hu / sd_mu * (mu - mu_0), sd_hu^2 * (1 - rho^2)),
# averaged over the conditional spread with the probit approximation to the logistic-normal
m1_pop = as_draws_df(m1) %>%
  as_tibble() %>%
  transmute(draw = .draw,
            mu_0 = b_Intercept,
            hu_0 = b_hu_Intercept,
            sd_mu = sd_replicate_group__Intercept,
            sd_hu = sd_replicate_group__hu_Intercept,
            rho = cor_replicate_group__Intercept__hu_Intercept)

detection_curve = expand_grid(m1_pop, log10_load = seq(1.5, 7, by = 0.1)) %>%
  mutate(hu_mean = hu_0 + rho * sd_hu / sd_mu * (log10_load * log(10) - mu_0),
         hu_var = sd_hu^2 * (1 - rho^2),
         p_detect = 1 - plogis(hu_mean / sqrt(1 + pi * hu_var / 8))) %>%
  group_by(log10_load) %>%
  summarise(med = median(p_detect),
            lo = quantile(p_detect, 0.05),
            hi = quantile(p_detect, 0.95),
            .groups = "drop")

p_det = ggplot() +
  geom_ribbon(data = detection_curve, aes(log10_load, ymin = lo, ymax = hi),
              fill = "grey70", alpha = 0.4) +
  geom_line(data = detection_curve, aes(log10_load, med), colour = "grey30") +
  geom_errorbar(data = m1_group_est,
                aes(y = p_detect, xmin = log10_load_lo, xmax = log10_load_hi, colour = detected_share),
                width = 0, alpha = 0.3) +
  geom_point(data = m1_group_est,
             aes(log10_load, p_detect, colour = detected_share, shape = qpcr_lab),
             alpha = 0.8) +
  labs(x = "latent log10(Bd load)", y = "P(detected)",
       colour = "detected", shape = "qPCR lab") +
  theme_bw() +
  theme(legend.position = "bottom")

# sigma is on the log scale of ln(load); convert to SD of log10(load)
m1_sigma = as_draws_df(m1) %>%
  as_tibble() %>%
  select(starts_with("b_sigma_noise_type")) %>%
  pivot_longer(everything(), names_to = "noise_type", values_to = "value") %>%
  mutate(noise_type = str_remove(noise_type, "b_sigma_noise_type"),
         sigma_log10 = exp(value) / log(10)) %>%
  group_by(noise_type) %>%
  summarise(med = median(sigma_log10),
            lo50 = quantile(sigma_log10, 0.25), hi50 = quantile(sigma_log10, 0.75),
            lo95 = quantile(sigma_log10, 0.025), hi95 = quantile(sigma_log10, 0.975),
            .groups = "drop")

p_sig = ggplot(m1_sigma, aes(noise_type, med)) +
  geom_linerange(aes(ymin = lo95, ymax = hi95)) +
  geom_linerange(aes(ymin = lo50, ymax = hi50), linewidth = 1.5) +
  geom_point(size = 2.5) +
  labs(x = "noise type", y = "noise SD (log10)") +
  theme_bw()

p_det + p_sig
```

### Posterior predictive checks

Share of results detected, and median positive log10 load, by replicate type × qPCR lab.

```{r m1-ppc}
#| code-fold: true
#| code-summary: "Show PPC code"
#| fig-width: 8
#| fig-height: 4

# pp_check only sees model variables, so call bayesplot directly with the grouping column
m1_yrep = posterior_predict(m1, ndraws = 500)

ppc_stat_grouped(y = m1_df$bd_load, yrep = m1_yrep, group = m1_df$ppc_group,
                 stat = function(y) mean(y > 0)) +
  labs(x = "share detected")

ppc_stat_grouped(y = m1_df$bd_load, yrep = m1_yrep, group = m1_df$ppc_group,
                 stat = function(y) median(log10(y[y > 0]))) +
  labs(x = "median log10(Bd load) among positives")
```

### Export for Model 2

Posterior draws and summaries (mean, SD, median, 95% interval) of population-level effects, group-level SDs, and correlations, saved to `fits/m1_replicates_export.rds` for use as priors in Model 2.

```{r m1-export}
#| code-fold: true
#| code-summary: "Show Export code"

m1_draws = as_draws_df(m1) %>%
  subset_draws(variable = c("^b_", "^sd_", "^cor_"), regex = TRUE)

m1_summary = m1_draws %>%
  summarise_draws(mean, sd, median,
                  ~ quantile(.x, probs = c(0.025, 0.975)))

m1_export = list(summary = m1_summary,
                 draws = m1_draws,
                 formula = m1_formula,
                 import_date = import_date,
                 export_date = Sys.Date())

saveRDS(m1_export, here(fit_dir, "m1_replicates_export.rds"))

m1_summary %>%
  kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```
