---
title: "rz_snarl_qpcr_comparison"
author: "Cob Staines"
format: html
editor: visual
---

# RZ-SNARL qPCR Comparison

**Objective**: Quantify diferences in Bd qPCR quantities between extraction and qPCR methods between RZ and SNARL labs.


## Setup
Load packages, set up working directory, establish database connection
```{r setup}
#| code-fold: true
#| code-summary: "Show Setup code"
#| results: hide
librarian::shelf(tidyverse, dbplyr, here, janitor, rio, ggplot2, knitr, kableExtra, RPostgres, DBI, RIBBiTR-BII/ribbitrrr)

wddir = here(Sys.getenv("data_dir"), "bd_qpcr_results", "rz_snarl_qpcr_comparison")

# establish database connection
dbcon = hopToDB("railway", hopReg = FALSE)

```

## Load data
- Pull in experiment data from data file: `qpcr_compare_data_260904.xlsx`
- Pull in replicate data from RIBBiTR database
```{r}
# load data
snarl_raw = import(here(wddir, "qpcr_compare_data_260904.xlsx"),
                   na = c("NA", "NaN"))

# replicates from database
db_bd = tbl(dbcon, Id("survey_data", "bd_qpcr_results"))
db_sample = tbl(dbcon, Id("survey_data", "sample"))
db_capture = tbl(dbcon, Id("survey_data", "capture"))

bd_replicates = db_bd %>%
  inner_join(db_sample, by = "sample_id") %>%
  inner_join(db_capture, by = "capture_id") %>%
  filter(ipc_pass | is.na(ipc_pass)) %>%
  group_by(sample_name_bd) %>%
  mutate(n_results = n()) %>%
  ungroup() %>%
  filter(standard_target_type == "ITS1",
         !is.na(bd_target_quant),
         !is.na(bd_its1_copies_per_swab),
         n_results > 1,
         qpcr_lab %in% c("snarl", "rz_lab")) %>%
  select(result_id,
         sample_id,
         capture_id,
         sample_name_bd,
         bd_cycle_quant,
         bd_target_quant,
         bd_its1_copies_per_swab,
         extraction_lab,
         qpcr_lab,
         extraction_kit,
         taxon_capture,
         comments_qpcr) %>%
  collect()
```

## Clean & Pivot
Clean column names, then pivot Bd load values wide by Frog ID
```{r clean}
#| code-fold: true
#| code-summary: "Show Clean & Pivot code"
#| 
snarl_clean = snarl_raw %>%
  clean_names() %>%
  rename(extraction_lab = lab_extract,
         extraction_method = extract_method,
         qpcr_lab = lab_qpcr,
         bd_quant_cycle = quant_cycle,
         bd_start_quant = start_quant,
         frog_id = frog_num) %>%
  mutate(extraction_lab = recode_values(extraction_lab,
                                        "pitt" ~ "rz",
                                        "snarl" ~ "snarl",
                                        default = NA),
         collect_lab = recode_values(collect_lab,
                                     "pitt" ~ "rz",
                                        "snarl" ~ "snarl",
                                        default = NA),
         qpcr_lab = recode_values(qpcr_lab,
                                     "pitt" ~ "rz",
                                        "snarl" ~ "snarl",
                                        default = NA),
         src = "snarl")

snarl_wide = snarl_clean %>%
  pivot_wider(id_cols = c("frog_id", "collect_lab"),
              names_from = c("extraction_method", "qpcr_lab"),
              values_from = "bd_load")

replicates_clean = bd_replicates %>%
  rename(bd_swab_id = sample_name_bd,
         bd_quant_cycle = bd_cycle_quant,
         bd_start_quant = bd_target_quant,
         bd_load = bd_its1_copies_per_swab,
         notes = comments_qpcr) %>%
  mutate(extraction_lab = recode_values(extraction_lab,
                                        "rz_lab" ~ "rz",
                                        default = extraction_lab),
         qpcr_lab = recode_values(qpcr_lab,
                                  "rz_lab" ~ "rz",
                                  default = qpcr_lab),
         extraction_method = recode_values(extraction_kit,
                                          "qiagen_dneasy" ~ "qiagen",
                                          default = extraction_kit),
         collect_lab = qpcr_lab,
         frog_id = bd_swab_id,
         src = "db",
         species_capture = str_to_sentence(gsub("_", " ", taxon_capture))) %>%
  select(all_of(colnames(snarl_clean))) %>%
  group_by(bd_swab_id) %>%
  mutate(replicate_id = row_number()) %>%
  ungroup()

replicates_wide = replicates_clean %>%
  pivot_wider(id_cols = c("frog_id", "collect_lab", "extraction_method", "qpcr_lab"),
              names_from = c("replicate_id"),
              values_from = "bd_load")

# snarl_wide_dat = snarl_wide %>%
#   select(-frog_id,
#          -collect_lab)

kable(snarl_wide, digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = TRUE,
                font_size = 12,
                fixed_thead = TRUE) %>%
  scroll_box(height = "400px")

```

## Data Summary
```{r summary}
#| code-fold: true
#| code-summary: "Show Data Summary code"
#| fig-width: 5
#| fig-height: 5
#| 
nice <- c(qiagen_rz = "Qiagen-RZ", qiagen_snarl = "Qiagen-SNARL",
          prepman_rz = "PrepMan-RZ", prepman_snarl = "PrepMan-SNARL")

# One row per panel; x/y orientation is set here
pairs <- tribble(
  ~letter, ~xvar,      ~yvar,
  "A",     "qiagen_rz",    "qiagen_snarl",
  "B",     "prepman_rz", "prepman_snarl",
  "C",     "qiagen_rz",    "prepman_rz",     # flipped from your version
  "D",     "qiagen_snarl", "prepman_snarl",
  "E",     "qiagen_rz",    "prepman_snarl"   # flipped from your version
)

# Stack all pairs into one long data frame
plot_df <- pmap_dfr(pairs, function(letter, xvar, yvar) {
  xv <- snarl_wide[[xvar]]
  yv <- snarl_wide[[yvar]]
  tibble(letter = letter, xvar = xvar, yvar = yvar, x = xv, y = yv)
}) %>%
  filter(!is.na(x), !is.na(y)) %>%
  mutate(xvar = factor(nice[xvar], levels = nice[1:3]),
         yvar = factor(nice[yvar], levels = nice[2:4]))

# Per-panel stats (on raw loads)
stats_df <- plot_df %>%
  group_by(letter, xvar, yvar) %>%
  summarise(n = n(), n_pos = sum(x > 0 & y > 0), .groups = "drop") %>%
  mutate(label = paste0("n = ", n, "\nboth > 0: ", n_pos))

ggplot(plot_df, aes(log10(x + 1), log10(y + 1))) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.6) +
  # geom_smooth(method = "lm") +
  geom_text(data = stats_df, aes(x = -Inf, y = Inf, label = letter),
            hjust = -0.6, vjust = 1.4, fontface = "bold", size = 5,
            inherit.aes = FALSE) +
  geom_text(data = stats_df, aes(x = Inf, y = -Inf, label = label),
            hjust = 1.05, vjust = -0.3, size = 2.75, lineheight = 0.9,
            inherit.aes = FALSE) +
  facet_grid(xvar ~ yvar, switch = "both") +
  coord_fixed() +
  labs(x = "log10(Bd load + 1)  [extraction method – qPCR lab]",
       y = "log10(Bd load + 1)  [extraction method – qPCR lab]") +
  theme_bw() +
  theme(strip.placement = "outside",
        strip.background = element_blank(),
        strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank())

```